Enforced Invariant Confluence

### 2020-01-14 doc/the_vldb_journal_paper/sections/interactive_decision_procedure.tex
Added tech report sections of paper.
Expanded correctness proof of decision proc.

    \algocomment{$R$ is a set of \sTIreachable{} states.}
    \algocomment{$NR$ is a set of \sTIunreachable{} states.}
    \algocomment{$I(s_0)$ is a precondition.}
    \Function{Helper}{$O$, $s_0$, $T$, $I$, $R$, $NR$}
      \State closed, $s_1$, $s_2$ $\gets$ \Call{IsIclosed}{$O$, $I - NR$}
      \If {closed}
        \Return true
      \EndIf
      \State Augment $R, NR$ with random search and user input
      \If{$s_1, s_2 \in R$}
        \Return false
      \EndIf
      \State \Return \Call{Helper}{$O$, $s_0$, $T$, $I$, $R$, $NR$}
    \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

\IsInvConfluent{} assumes access to an \invariantclosure{} decision procedure
\IsIclosed$(O, I)$. \IsIclosed$(O, I)$ returns a triple (closed, $s_1$, $s_2$).
closed is a boolean indicating whether $O$ is \Iclosed{}. If closed is true,
then $s_1$ and $s_2$ are null. If closed is false, then $s_1$ and $s_2$ are a
counterexample witnessing the fact that $O$ is not \Iclosed{}. That is,
$I(s_1)$ and $I(s_2)$, but $\lnot I(s_1 \join s_2)$ (e.g., $s_1$ and $s_2$ from
\exampleref{Z2}).  As we mentioned earlier, we can (and do) implement the
\invariantclosure{} decision procedure using an SMT solver like
Z3~\cite{de2008z3}.

\IsInvConfluent{} first checks that $s_0$ satisfies the invariant. $s_0$ is
reachable, so if it does not satisfy the invariant, then $O$ is not
\sTIconfluent{} and \IsInvConfluent{} returns false. Otherwise,
\IsInvConfluent{} calls a helper function \Helper{} that---in addition to $O$,
$s_0$, $T$, and $I$---takes as arguments a set $R$ of \sTIreachable{} states
and a set $NR$ of \sTIunreachable{} states. Like \IsInvConfluent, \Helper$(O,
s_0, T, I, R, NR)$ returns whether $O$ is \sTIconfluent{} (assuming $R$ and
$NR$ are correct).  As \algoref{InteractiveDecisionProcedure} executes, $NR$ is
iteratively increased, which removes unreachable states from $I$ until $I$ is a
subset of $\setst{s \in S}{\sTIreachablepredicate{s}}$.

First, \Helper{} checks to see if $O$ is \iclosed{(I - NR)}.
%
If \IsIclosed{} determines that $O$ is \iclosed{(I - NR)}, then by
\thmref{IclosureImpliesIconfluence}, $O$ is \sticonfluent{s_0}{T}{I-NR}, so
\[
  \setst{s \in S}{\stireachablepredicate{s_0}{T}{I-NR}{s}}
    \subseteq I - NR
    \subseteq I
\]
Because $NR$ only contains \sTIunreachable{} states, then the set of
\sTIreachable{} states is equal to set of \stireachable{s_0}{T}{I-NR}{} states
which, as we just showed, is a subset of $I$.  Thus, $O$ is \sTIconfluent{}, so
\Helper{} returns true.

If \IsIclosed{} determines that $O$ is \emph{not} \iclosed{(I - NR)}, then we
have a counterexample $s_1, s_2$. We want to determine whether $s_1$ and $s_2$
are reachable or unreachable. We can do so in two ways.
%
First, we can randomly generate a set of reachable states and add them to $R$.
If $s_1$ or $s_2$ is in $R$, then we know they are reachable.
%
Second, we can prompt the user to tell us directly whether or not the states
are reachable or unreachable.

@@

After $s_1$ and $s_2$ have been labelled as \sTIreachable{} or not, we
continue. If both $s_1$ and $s_2$ are \sTIreachable{}, then so is $s_1 \join
s_2$, but $\lnot I(s_1 \join s_2)$. Thus, $O$ is not \sTIconfluent{}, so
\Helper{} returns false. Otherwise, one of $s_1$ and $s_2$ is
\sTIunreachable{}, so we recurse.

@@

\subsection{Limitations}
Our interactive \invariantconfluence{} decision procedure has two
limitations. First, \algoref{InteractiveDecisionProcedure} requires an
\invariantclosure{} decision procedure, but determining \invariantclosure{} is
undecidable in general.
%
\begin{techreport}

  For example, let $O_p = (S, \join)$ where $S$ is the set of all programs and
  $s_1 \join s_2 = p$ for some fixed program $p$. Letting $I$ be the set of all
  programs that terminate, determining if $O_p$ is \Iclosed{} is tantamount to
  determining if $p$ terminates.
\end{techreport}
%
In practice, we can implement an \invariantclosure{} decision procedure using
an SMT solver like Z3 that works well on simple objects, invariants, and merge
operators (e.g., integers, tuples, infinite sets, bit vectors, linear
constraints, basic arithmetic, tuple projection, basic set operations, bit
arithmetic). But, SMT solvers are mostly unable to analyze more complex
constructs (e.g., finite lists~\cite{kroning2009proposal}, graphs, recursive
algebraic data types, nonlinear constraints, merge operators that contain loops
or recursion).

@@

\begin{techreport}
  \subsection{Tolerating User Error}
  \algoref{InteractiveDecisionProcedure} is an \emph{interactive} decision
  procedure. It requires that a user classify states as reachable or
  unreachable.  To err is human, so what happens when a user incorrectly
  classifies a state?  There are two possible errors that can be made, and
  \algoref{InteractiveDecisionProcedure} can be made robust to both.

@@

  \textbf{A user can label a reachable state as unreachable.}
  In this case, \IsIclosed$(O, I - NR)$ might return true, even though $O$ is
  not \sTIconfluent. Thus, a user might falsely believe their distributed
  object to be \sTIconfluent{} even though it isn't, and eventually one replica
  of their distributed object might enter a state that violates the invariant.
  %
  We can mitigate this in two ways. First, we can aggressively expand $R$
  automatically. If a user ever labels a state $s$ as unreachable, but $s \in
  R$, we can notify the user of their mistake. Second, \Helper{} returns true
  when $O$ is \iclosed{(I - NR)} for some $NR$, so $O$ \emph{is}
  \sticonfluent{s_0}{T}{I-NR} (even though it might not be \sTIconfluent{}).
  Thus, when a user replicates their distributed object across a set of
  servers, they can deploy with the invariant $I - NR$ instead of $I$. If $NR$
  is correct and only contains unreachable states, then deploying with $I - NR$
  is equivalent to deploying with $I$. If $NR$ is incorrect and contains some
  \sTIreachable{} states, then some of these states are artificially made
  unreachable, but the system is still guaranteed to never produce a state
  that violates $I$.
\end{techreport}

>>>>

    \algocomment{$R$ is a set of \sTIreachable{} states.}
    \algocomment{$NR$ is a set of \sTIunreachable{} states.}
    \algocomment{$I(s_0)$ is a precondition.}
    \Function{Helper}{$O$, $s_0$, $T$, $I$, $R$, $NR$}
      \State closed, $s_1$, $s_2$ $\gets$ \Call{IsIclosed}{$O$, $I - NR$}
      \If {closed}
        \State \Return true
      \EndIf
      \State Augment $R, NR$ with random search and user input
      \If{$s_1, s_2 \in R$}
        \State \Return false
      \EndIf
      \State \Return \Call{Helper}{$O$, $s_0$, $T$, $I$, $R$, $NR$}
    \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

\IsInvConfluent{} assumes access to an \invariantclosure{} decision procedure
\IsIclosed$(O, I)$. The decision procedure \IsIclosed$(O, I)$ returns a triple
(closed, $s_1$, $s_2$).  closed is a boolean indicating whether $O$ is
\Iclosed{}. If closed is true, then $s_1$ and $s_2$ are null. If closed is
false, then $s_1$ and $s_2$ are a counterexample witnessing the fact that $O$
is not \Iclosed{}. That is, $I(s_1)$ and $I(s_2)$, but $\lnot I(s_1 \join s_2)$
(e.g., $s_1$ and $s_2$ from \exampleref{Z2}).  As we mentioned earlier, we can
(and do) implement the \invariantclosure{} decision procedure using an SMT
solver like Z3~\cite{de2008z3}.

\IsInvConfluent{} first checks that $s_0$ satisfies the invariant. $s_0$ is
reachable, so if it does not satisfy the invariant, then $O$ is not
\sTIconfluent{} and \IsInvConfluent{} returns false. Otherwise,
\IsInvConfluent{} calls a helper function \Helper{} that---in addition to $O$,
$s_0$, $T$, and $I$---takes as arguments a set $R$ of \sTIreachable{} states
and a set $NR$ of \sTIunreachable{} states. Like \IsInvConfluent, the function
\Helper$(O, s_0, T, I, R, NR)$ returns whether $O$ is \sTIconfluent{} (assuming
$R$ and $NR$ are correct).  As \algoref{InteractiveDecisionProcedure} executes,
$NR$ is iteratively increased, which removes unreachable states from $I$ until
$I$ is a subset of $\setst{s \in S}{\sTIreachablepredicate{s}}$.

First, \Helper{} checks to see if $O$ is \iclosed{(I - NR)}.
%
If \IsIclosed{} determines that $O$ is \iclosed{(I - NR)}, then by
\thmref{IclosureImpliesIconfluence}, $O$ is \sticonfluent{s_0}{T}{I-NR}, so
\[
  \hfill
  \setst{s \in S}{\stireachablepredicate{s_0}{T}{I-NR}{s}}
    \subseteq I - NR
    \subseteq I
  \hfill
\]

In this case, $O$ is \sTIconfluent{} because the set of \sTIreachable{} states
is a subset of \stireachable{s_0}{T}{I-NR}{} states which is in turn a subset
of $I$. We prove that now.
\begin{proof}
  We prove, by structural induction, that every \sTIreachable{} expression $e$
  is also \stireachable{s_0}{T}{I-NR}{}.

  \begin{itemize}
    \item \textbf{Case $s_0$.}
      $s_0$ is \stireachable{s_0}{T}{I-NR}{} by definition.

    \item \textbf{Case $t(e)$.}
      $e$ is \sTIreachable{}, so by the inductive hypothesis, it is also
      \stireachable{s_0}{T}{I-NR}{}. Moreover, $t(e) \in I$ and $t(e) \notin
      NR$ (by assumption), so $t(e) \in I - NR$. Thus, $t(e)$ is
      \stireachable{s_0}{T}{I-NR}{}.

    \item \textbf{Case $e_1 \join e_2$.}
      By the inductive hypothesis, $e_1$ and $e_2$ are both
      \stireachable{s_0}{T}{I-NR}{}, so $e_1 \join e_2$ is also
      \stireachable{s_0}{T}{I-NR}{}.
  \end{itemize}
\end{proof}

If \IsIclosed{} determines that $O$ is \emph{not} \iclosed{(I - NR)}, then we
have a counterexample $s_1, s_2$. That is, $s_1, s_2 \in I-NR$, but $s_1 \join
s_2 \notin I-NR$. We want to determine whether $s_1$ and $s_2$ are
\sTIreachable{} or unreachable. We can do so in two ways.
%
First, we can randomly generate a set of reachable states and add them to $R$.
If $s_1$ or $s_2$ is in $R$, then we know they are reachable.
%
Second, we can prompt the user to tell us directly whether or not the states
are reachable or unreachable.

@@

After $s_1$ and $s_2$ have been labelled as \sTIreachable{} or not, we
continue. If both $s_1$ and $s_2$ are \sTIreachable{}, then so is $s_1 \join
s_2$, but $\lnot I(s_1 \join s_2)$ (because $s_1 \join s_2 \notin I - NR$).
Thus, $O$ is not \sTIconfluent{}, so \Helper{} returns false. Otherwise, one of
$s_1$ and $s_2$ is \sTIunreachable{}, so we recurse.

@@

\subsection{Limitations}
Our interactive \invariantconfluence{} decision procedure has two
limitations. First, \algoref{InteractiveDecisionProcedure} requires an
\invariantclosure{} decision procedure, but determining \invariantclosure{} is
undecidable in general.
%
For example, let $O_p = (S, \join)$ where $S$ is the set of all programs and
$s_1 \join s_2 = p$ for some fixed program $p$. Letting $I$ be the set of all
programs that terminate, determining if $O_p$ is \Iclosed{} is tantamount to
determining if $p$ terminates.
%
In practice, we can implement an \invariantclosure{} decision procedure using
an SMT solver like Z3 that works well on simple objects, invariants, and merge
operators (e.g., integers, tuples, infinite sets, bit vectors, linear
constraints, basic arithmetic, tuple projection, basic set operations, bit
arithmetic). But, SMT solvers are mostly unable to analyze more complex
constructs (e.g., finite lists~\cite{kroning2009proposal}, graphs, recursive
algebraic data types, nonlinear constraints, merge operators that contain loops
or recursion).

@@

\subsection{Tolerating User Error}
\algoref{InteractiveDecisionProcedure} is an \emph{interactive} decision
procedure. It requires that a user classify states as reachable or
unreachable.  To err is human, so what happens when a user incorrectly
classifies a state?  There are two possible errors that can be made, and
\algoref{InteractiveDecisionProcedure} can be made robust to both.

@@

\textbf{A user can label a reachable state as unreachable.}
In this case, \IsIclosed$(O, I - NR)$ might return true, even though $O$ is
not \sTIconfluent. Thus, a user might falsely believe their distributed
object to be \sTIconfluent{} even though it isn't, and eventually one replica
of their distributed object might enter a state that violates the invariant.
%
We can mitigate this in two ways. First, we can aggressively expand $R$
automatically. If a user ever labels a state $s$ as unreachable, but $s \in
R$, we can notify the user of their mistake. Second, \Helper{} returns true
when $O$ is \iclosed{(I - NR)} for some $NR$, so $O$ \emph{is}
\sticonfluent{s_0}{T}{I-NR} (even though it might not be \sTIconfluent{}).
Thus, when a user replicates their distributed object across a set of
servers, they can deploy with the invariant $I - NR$ instead of $I$. If $NR$
is correct and only contains unreachable states, then deploying with $I - NR$
is equivalent to deploying with $I$. If $NR$ is incorrect and contains some
\sTIreachable{} states, then some of these states are artificially made
unreachable, but the system is still guaranteed to never produce a state
that violates $I$.
